### 2013-05-20 Publications/2013-TMI/03-introduction.tex
Drafted validation
I wrote renewed Data and methods section for the TMI version.

As long as this summarizes the issue, close #59. Some references
have been fixed regarding #55.

Added some people to the acks.

We suggest clustering the current methodologies of template-based segmentation 
methods into three groups. The first group typically adds a shape prior term to 
the energy functional of an evolving active contour~\citep{Bresson2006a,Chan2005,
Chen2002,Cremers2006,Gastaud2004,Paragios2003,Vemuri2003a,Yezzi2003a}.
These methods generally have a explicit description of the expected relative boundary 
locations of the object to be delineated, and some even model the statistical deviations
from this average shape. Closely related to this group are atlas-based segmentation
methods \citep{Gorthi2011,Gorthi2009,Pohl2005,Pohl2006,Wang2006}, where the prior 
imposes consistent voxel-based classification of contiguous regions. Here, the 
presence of more structures than one unique \gls{roi} helps aligning the target image 
with the atlas in a hierarchical fashion. Finally, the third group generalizes 
the atlas to actual images, and the contour is to segment simultaneously two 
different target images, related by a spatial transform to be co-estimated~\citep{Wyatt2003,
Yezzi2003}.

>>>>

We suggest clustering the current methodologies of template-based segmentation 
methods into three groups. The first group typically adds a shape prior term to 
the energy functional of an evolving active contour
\citep{bresson_variational_2006,chan_level_2005,chen_using_2002,cremers_kernel_2006,
gastaud_combining_2004,paragios_level_2003,vemuri_joint_2003,yezzi_variational_2003}.
These methods generally have a explicit description of the expected relative boundary 
locations of the object to be delineated, and some even model the statistical deviations
from this average shape. Closely related to this group are atlas-based segmentation
methods \citep{gorthi_segmentation_2009,gorthi_active_2011,pohl_unifying_2005,
pohl_bayesian_2006,wang_joint_2006}, where the prior 
imposes consistent voxel-based classification of contiguous regions. Here, the 
presence of more structures than one unique \gls{roi} helps aligning the target image 
with the atlas in a hierarchical fashion. Finally, the third group generalizes 
the atlas to actual images, and the contour is to segment simultaneously two 
different target images, related by a spatial transform to be co-estimated
\citep{wyatt_map_2003,yezzi_variational_2003}.

### 2013-05-20 Publications/2013-TMI/05-experiments.tex
Drafted validation
I wrote renewed Data and methods section for the TMI version.

As long as this summarizes the issue, close #59. Some references
have been fixed regarding #55.

Added some people to the acks.

\section{Data and experiments}
\label{sec:experiments}
%
\subsection{Shape prior}
%
As described in \autoref{sec:introduction}, the general situation in
the connectivity pipelines consists of having 
a reliable segmentation obtained from the high resolution \gls{t1} 
reference image. Therefore, a precise location of the tissue interfaces
of interest is available in a reference space. Given that the anatomical 
reference segmentation is beyond the scope of this manuscript, we simply 
rely on the true contours known from the underlying models, and do not 
seek to establish them in a separate segmentation step on ``anatomical'' images.
%
\subsection{Synthetic gray-scale data}
%
The first simplified model to test the approach is inspired by a problem 
shown for coupled \gls{csf}/\gls{gm} and \gls{gm}/\gls{wm} segmentation in \citep{macdonald_automated_2000}. Authors note that ``partial volume effects 
blur the distinction between closely adjacent surfaces in deep sulci, leading 
to a well-known segmentation error in which the deeper reaches of sulci are 
not penetrated by the putative surface model.'' This problem is aggravated 
in DWI, since the resolution tends to be worse compared to the anatomical 
images considered in \citep{macdonald_automated_2000}. They test their 
coupled segmentation algorithm on an image, ``representing a sulcus in 
which the distinction between opposing banks of the sulcus has been obscured 
by partial volume.''

Here, we reproduce their model on a volume consisting of three 
piecewise-constant parts: a notched ball representing the \gls{wm} with a single 
sulcus ($\mu_{WM} = 0.8$), a cortical sheet of \gls{gm} obtained through dilation 
of the \gls{wm} ($\mu_{GM} = 0.5$), and the surrounding background representing 
\gls{csf} ($\mu_{CSF} = 0.2$). The volume is then affected by additive Gaussian 
noise, effectively creating uniform standard deviation of $\sigma = 0.045$ per 
region.

As illustrated in \autoref{fig:sulcusmodel}, conventional single surface 
segmentation of the \gls{csf}/\gls{gm} boundary misses to capture the sulcus in 
its full depth. With our proposed model, we expect the joint 
segmentation-registration to be driven largely by the inner, \gls{gm}/\gls{wm}
contour that exhibits sufficient contrast and lesser partial volume effects. 
The shape prior of the outer, difficult contour will then be co-aligned through 
the regularity of the estimated deformation field.

\begin{figure}
\begin{tabular}{ccccc}
\includegraphics[width=0.19\textwidth]{contours_wrong} & \includegraphics[width=0.19\textwidth]{contours_inner} &
\includegraphics[width=0.19\textwidth]{contours_both} &
\includegraphics[width=0.19\textwidth]{pialsurf} &
\includegraphics[width=0.19\textwidth]{gmwmsurf}\\
a)&b)&c)&d)&e)
\end{tabular}
\caption{The gray-scale sulcus model. a) The apparent CSF/GM boundary is affected by partial volume in the sulcal cavity, and conventional segmentation is likely to miss it. b) The GM/WM interface here has consistently good contrast. c) Registering the two shape priors coupled through deformation field regularity is expected to guide the CSF/GM contour. d\&e) 3D view of the two shape priors.}
\label{fig:sulcusmodel}
\end{figure}

%
\subsection{Simulated diffusion data}
%
In order to demonstrate the functionality of the methodology, 
and characterize its possibilities with diffusion data,
we built a synthetic phantom from a model consisting of several 
spherical shapes emulating the different brain tissues (see 
\autoref{fig:fa}, first row). We simulated a single-shell
acquisition at $b=1000$ \eqref{eqn:MultiTensor} with 30 samples
equally distributed on one hemisphere, corresponding to a standard
\gls{dti} acquisition commonly used in clinical practice.
We generated a synthetic displacement field to produce
an \gls{epi}-like distortion on the \gls{dwi} signal. Finally,
we reconstructed the deformed signal, and generated \gls{dwi}-derived
scalar maps (specifically the \gls{fa} and \gls{md} maps).
Then, the scalar maps were placed as features in 
\eqref{eq:complete_energy} to drive our segmentation approach, 
finding the location of the \gls{wm}-\gls{gm}
and the \gls{csf}-\gls{wm} interfaces in the distorted space. \\

\paragraph{Signal simulation}
To numerically simulate the \gls{mri} signal attenuation when applying a diffusion 
gradient in a voxel with $N$ fiber populations we made use of the standard 
\emph{Multi-Tensor Model}~\cite{Tuch:2002aa}:
%
\begin{equation} 
\label{eqn:MultiTensor}
S(q) / S_{0} = \sum_{i=1}^{N} f_{i} \, \exp{ \left( -b \, q^{T} \, \mathbf{D}_i \, q\right) } + f_{iso} \, \exp{ \left( -b \, \mathbf{D}_{iso} \right) }  + f_{gm} \, \exp{ \left( -b \, \mathbf{D}_{gm} \right) } ,
\end{equation}
%
where $q \in \mathbb{S}^2$ is the direction of the diffusion gradient applied,
$b$ is the b-value accounting for its strength and $S_0 \equiv S(0)$ is the 
signal with no diffusion weighting. $f_i$ and $\mathbf{D}_i$ are the volume 
fraction and the diffusion tensor characterizing the $i$-th fiber population, 
whereas the tensors $\mathbf{D}_{iso}$ and $\mathbf{D}_{gm}$ describe the 
diffusion processes of partial volumes with \gls{csf} and \gls{gm} within the 
voxel, which volume fractions are, respectively, $f_{iso}$ and $f_{gm}$, and
$\sum_i f_i + f_{iso} + f_{gm} = 1$. In this work, the diffusion properties have
been taken from standard ranges typically observed in the human 
brain~\cite{Canales-Rodriguez:2009aa}. Moreover we assumed $N = 1$ and $S_0 = 1$ 
without loss of generality.

\paragraph{Noise simulation}
The diffusion MRI signal $S$ has been corrupted with 
\textit{Rician noise}~\cite{Gudbjartsson:1995aa} as follows:
%
\begin{equation}
	\tilde{S} = \sqrt{ (S + \varepsilon_1)^2 + \varepsilon_2^2 }
\end{equation}
%
where $\varepsilon_{1,2}$ are Gaussian distributed with zero mean
and standard deviation $\sigma = S_0 / \mathit{SNR}$ and \gls{snr}
is the \gls{snr} on the $S_0$ image.

\paragraph{Simulated \gls{epi} distortion}
For this model, we created manually a sound distortion to approximate
the real \gls{epi} distortions. We interpolated the distortion to a 
dense deformation field, and applied it to generate the deformed \gls{dwi}
signal.

\paragraph{Derived scalar features}
We obtained the local fiber configuration in each voxel with a commonly 
used \gls{dti} reconstruction tool~\footnote{DTIFIT, included
in the FMRIB's Software Library (FSL), 
\url{http://fsl.fmrib.ox.ac.uk/fsl/fsl-4.1.9/fdt/fdt_dtifit.html}}. 
The properties of the reconstructed tensors and derived scalar maps have
been studied by \cite{ennis_orthogonal_2006}. Based on their
findings, \gls{fa}~\eqref{eq:fa} and \gls{md}~\eqref{eq:md} are
considered complementary features, and therefore we selected them for the 
energy model \eqref{eq:complete_energy} in driving the 
registration-segmentation process. 
Whereas \gls{fa} informs mainly about the \emph{shape} of diffusion, 
the \gls{md} is more related to the \emph{magnitude} of the process:

\begin{align}
\mathrm{FA} &= \sqrt{ \frac{1}{2}}\,\frac{\sqrt{ (\lambda_1 - \lambda_2)^2 + (\lambda_2 - \lambda_3)^2 + (\lambda_3 - \lambda_1)^2}}{\sqrt{ {\lambda_1}^2 + {\lambda_2}^2 + {\lambda_3}^2}} \label{eq:fa} \\
\mathrm{MD} &= ( \lambda_1 + \lambda_2 + \lambda_3 ) / 3 \label{eq:md}
\end{align}
where $\lambda_i$ are the eigenvalues of the diffusion tensor 
associated with the diffusion signal $S(\vec{q})$. There exist 
two main reasons to justify their choice. 
First, they are well-understood and standardized in clinical routine.
Second, together they contain most of the information that is
usually extracted from the \gls{dwi}-derived scalar maps
\cite{ennis_orthogonal_2006}. The different region properties 
$(\mathbf{\mu}_R,\Sigma_R)$ measured for the \gls{fa} and \gls{md}
joint distribution are summarized in \autoref{table:parameters}. \\

\begin{table}
\caption{Model means and covariances of \gls{fa} and \gls{md} estimated 
from the reconstructed simulated \gls{dwi} images for each modeled tissue, 
\gls{wm}, \gls{gm}, and \gls{csf}. As expected, the two scalar features 
are complementary and the three tissues can well be discriminated. }
\label{table:parameters}
%\begin{tabularx}{1.0\textwidth}{c|XXX}
\begin{tabular}{*{3}{@{}p{0.2\textwidth}}@{}p{0.4\textwidth}@{}}
\toprule
% tissue        & $\mathbf{\mu}_{FA}$ & $\mathbf{\mu}_{MD}$ & $\Sigma$ \newline
        & $\mathbf{\mu}$ \\
\cmidrule(r){2-3}
Tissue& FA & MD &$\Sigma$\\
\cmidrule(r){1-1}\cmidrule(r){2-2}\cmidrule(r){3-3}\cmidrule{4-4}
\gls{wm}  & 0.778 & 6.94\e{-4} & 
   $\begin{pmatrix}
   	4.85\e{-3} & -6.90\e{-6} \\ -6.90\e{-6} & 1.03\e{-8}
   \end{pmatrix}$\vspace{2mm}\\
%
\gls{gm}  & 0.119 & 8.95\e{-4} &
   $\begin{pmatrix}
   	5.90\e{-4} & -1.43\e{-6} \\ -1.43\e{-6} & 1.04\e{-8}
   \end{pmatrix}$\vspace{2mm}
\\
%
\gls{csf} & 0.103 & 2.99\e{-3} &
   $\begin{pmatrix}
   	1.19\e{-3} & 2.22\e{-7} \\ 2.22\e{-7} & 1.56\e{-8}
   \end{pmatrix}$
\\
\bottomrule
\end{tabular}
%\end{tabularx}

\end{table}

>>>>

\section{Data and experiments}
\label{sec:experiments}
%

\subsection{Image Data}
\label{sec:datasets}

\subsubsection{Simulated digital phantom} %
%
The lack of a widely accepted \emph{gold}-standard in the application
field has been addressed by several authors \citep{cote_tractometer:_2013}
In this work, we use one of the most complete, and publicly available,
digital phantoms (\url{http://hardi.epfl.ch/static/events/2013_ISBI/testing_data.html}).
The phantom is a spherical volume containing a set of fiber bundles, that connect 
one area of a ``cortex'' to another. The model accounts for \gls{pve} using 
a similar approach to \citep{close_software_2009} and for \gls{csf} contamination
as well.

\subsubsection{Real \gls{mri} datasets} %
%
We used image data from XX healthy volunteers with no history of neurological
conditions (ages XX$\pm$XX, X female) to illustrate the applicability of our 
approach. All the subjects were scanned in a 3T MR Scanner (Siemens Magnetom
TrioTim blah blah)
with an X-channel head coil. Subjects were scanned twice with the same protocol,
described hereafter. After being scanned the first time, each subject exited the
scan room for a short break and then reentered for an identical scan session.
To note, there was a full repositioning of the volunteer, coils, blankets and pads
before each scan and re-scan session.

The scan session protocol was as follows:
\begin{enumerate}
\item Triplanar survey (Localizer).
\item Field Mapping: field mapping using GRE sequence was performed before \gls{dwi}
acquisition for susceptibility correction purposes.
\item DTI: \gls{dwi} were acquired with axial in-plane 
isotropic resolution $2mm$, slice thickness $2mm$, $XXX \times XXX \times XXX$ 
image matrix, TR= XXXX ms, TE = XX ms, NEX = X, BW = XXXX Hz/pixel, GRAPPA
acceleration factor 2. The series included images acquired with diffusion
weighting along 30 non-collinear directions ( $b$=700sm$^{-2}$ ) sampled
twice for averaging, and 5 interleaved images acquired without diffusion 
weighting ( $b=0$ ).
\item Field Mapping (same as before DTI)
\item Structural T1: An MPRAGE T1-weighted acquisition, sagittal
GRE sequence, in-plane isotropic resolution 1.0 mm, slice thickness 1.2mm
$XXX \times XXX \times XXX$ image matrix, TR=2300 ms, TE=2.98 ms, FA= 9, NEX=X, BW= 240 Hz/pixel.
\item Structural T2: A T2-weighted acquisition, oblique axial
TSE sequence, in-plane isotropic resolution 1.0 mm, slice thickness 1.2mm
$XXX \times XXX \times XXX$ image matrix, TR=3200 ms, TE= 408 ms, NEX=X, BW= 751 Hz/pixel.
\end{enumerate}

\subsection{Image preprocessing and shape-prior generation}
\label{sec:preprocessing}
%
Regardless the dataset type (simulated or real), all \gls{dwi} datasets were processed
using the standard \gls{dti} reconstruction methods provided by FSL \footnote{DTIFIT, 
included in the FMRIB's Software Library (FSL), 
\url{http://fsl.fmrib.ox.ac.uk/fsl/fsl-4.1.9/fdt/fdt_dtifit.html}} to fit
the tensor model and produce scalar maps of the required features.
The properties of the reconstructed tensors and derived scalar maps have
been studied by \cite{ennis_orthogonal_2006}. Based on their
findings, \gls{fa}~\eqref{eq:fa} and \gls{md}~\eqref{eq:md} are
considered complementary features, and therefore we selected them for the 
energy model \eqref{eq:complete_energy} in driving the 
registration-segmentation process. 
Whereas \gls{fa} informs mainly about the \emph{shape} of diffusion, 
the \gls{md} is more related to the \emph{magnitude} of the process:

\begin{align}
\mathrm{FA} &= \sqrt{ \frac{1}{2}}\,\frac{\sqrt{ (\lambda_1 - \lambda_2)^2 + (\lambda_2 - \lambda_3)^2 + (\lambda_3 - \lambda_1)^2}}{\sqrt{ {\lambda_1}^2 + {\lambda_2}^2 + {\lambda_3}^2}} \label{eq:fa} \\
\mathrm{MD} &= ( \lambda_1 + \lambda_2 + \lambda_3 ) / 3 \label{eq:md}
\end{align}
where $\lambda_i$ are the eigenvalues of the diffusion tensor 
associated with the diffusion signal $S(\vec{q})$. There exist 
two main reasons to justify their choice. 
First, they are well-understood and standardized in clinical routine.
Second, together they contain most of the information that is
usually extracted from the \gls{dwi}-derived scalar maps
\cite{ennis_orthogonal_2006}.

Preprocessing differed between simulated and real datasets in the 
shape-prior surfaces generation, as we describe hereafter.

\subsubsection{Simulated digital phantom} %
A description goes here: surfaces extraction, synthetic susceptibility artifact generation.

\subsubsection {Real \gls{mri} data} %

We used a standard automated method 
available in \emph{FreeSurfer} \citep{fischl_freesurfer_2012} to obtain the
cortical gray/white boundary from the T1 scan \citep{greve_accurate_2009}.
Parcellations used to evaluate the repeatability of the method are also
obtained from \emph{FreeSurfer}.

\subsection{Experiments and evaluation}
\label{sec:experiments_evaluation}
%
\subsubsection{Validation on the simulated digital phantom} %
\label{sec:validation_phantom}
We firstly evaluated our approach on the simulated data, using
the distorted data with a deformation field similar to the
susceptibility artifact that affects real \gls{dwi} data. To 
this end, we report the following indices:
\begin{itemize}
\item \gls{se}, a distance between the one-to-one corresponding
vertices, weighted by their respective Voronoi area.
\begin{equation}
MSE = \sum\limits_k \sum\limits_j^M w_j\,\| \mathbf{x}_j - \hat{\mathbf{x}}_j \|
\end{equation}
where $\mathbf{x}_j$ are the locations of the $M$ vertices of the $k$ priors,
$\hat{\mathbf{x}}_j$ are the corresponding locations recovered, and $w_j$ the
weighting factor as the relative surface of the Voronoi area.
\item \gls{wi}, L2-distance between the theoretical and the recovered
deformation field.
\begin{equation}
WI = \frac{1}{N} \sum\limits_i^N \| \mathbf{d}_i - \hat{\mathbf{d}}_i \|
\end{equation}
where $\mathbf{d}_i$ is the theoretical displacement vector at position $i$
and $\hat{\mathbf{d}}_i$ is the recovered one at the same index position.
\item Parcellation agreement, the \gls{se} averaged by defined \glspl{roi}
between the theoretical and the recovered parcellations.
\item \gls{nof} agreement, between the fibers recovered on the original
data and the processed data.
\end{itemize}

Additionally, the entropy of data is studied in both
original and recovered datasets to draw another possible basis for the 
assessment of the real datasets. We also report the same indices for the
outcome of a widely-used methodology that combines field-map susceptibility 
correction and T1-T2-\gls{dwi} registration (\textbf{FIXME}: I suggest to give 
it a compact name and define it in the pre-processing section or even before
in the intro).

\subsubsection{Evaluation on real datasets}
For the real datasets there is no published \emph{gold}-standard to validate results.
Thus, visual results are provided to let compare the performance with the NICEACRONYM
standard methodology. Additionally, cross-comparison of repeatability results are 
provided. In this second evaluation strategy, all the indices defined in 
\autoref{sec:validation_phantom} are reported.

### 2013-05-20 Publications/2013-TMI/06-results.tex
Drafted validation
I wrote renewed Data and methods section for the TMI version.

As long as this summarizes the issue, close #59. Some references
have been fixed regarding #55.

Added some people to the acks.

\subsection{Synthetic gray-scale data}
The proposed method solved precisely the specific challenge
created by the model. A severe distortion of the model is
artificially created adding complexity to the problem of
partial voluming in the outer contour of the sulcus.
\autoref{fig:sulcusmodel_result} provides visual assessment
for this result. With 16x16x16 control points, 
computation time for this model was around
10 minutes in a {\color{red}{(specific machine details)}}.

\begin{figure}
\begin{tabular}{ccccc}
\includegraphics[width=0.19\textwidth]{model2result_b_1} &
\includegraphics[width=0.19\textwidth]{model2result_b_2} &
\includegraphics[width=0.19\textwidth]{model2result_a_1} &
\includegraphics[width=0.19\textwidth]{model2result_a_2} &
\includegraphics[width=0.19\textwidth]{model2surf} \\
\multicolumn{2}{c}{a)} & \multicolumn{2}{c}{b)} & c)
\end{tabular}
\caption{Result of the segmentation performance on the sulcus model.
a) Two slices and prior contours before segmentation-registration. b) The same slices and contours after distortion estimation. c) 3D rendering of the deformed priors, with the traces of the two selected slices highlighted.}
\label{fig:sulcusmodel_result}
\end{figure}

\subsection{Simulated diffusion data}
%
The proposed method successfully reverted the synthetic distortion
field we applied to the data. With 16x16x16 control points, the
displacements field is dense enough to correctly represent the
synthetic field. Second row in \autoref{fig:fa} shows the fitted
contours obtained by using the original surfaces of the model
as shape priors, with a constant translation of (5.0mm,10mm.,-5.0mm)
to illustrate briefly the extent of the capture range of the algorithm.
Computation time in this case was around 4 minutes in the previously
described platform. \\
%

\begin{figure}
\begin{tabular}{ccccc}
\includegraphics[width=0.19\textwidth]{model1result_b_1} &
\includegraphics[width=0.19\textwidth]{model1result_b_2} &
\includegraphics[width=0.19\textwidth]{model1result_b_3} &
\includegraphics[width=0.19\textwidth]{model1result_b_4} &
\includegraphics[width=0.19\textwidth]{model1result_b_5} \\
\includegraphics[width=0.19\textwidth]{model1result_a_1} &
\includegraphics[width=0.19\textwidth]{model1result_a_2} &
\includegraphics[width=0.19\textwidth]{model1result_a_3} &
\includegraphics[width=0.19\textwidth]{model1result_a_4} &
\includegraphics[width=0.19\textwidth]{model1result_a_5} \\
\end{tabular}
\caption{First row presents several slices along Z axis of the distorted \gls{fa} map and
the undistorted \gls{wm}-\gls{gm} and \gls{wm}-\gls{csf} contours given as shape priors. Second row contains the same map, now with the contours after joint segmentation-registration.}
\label{fig:fa}
\end{figure}

>>>>

Please! Write me ASAP.
